A thermodynamic view of dusty protoplanetary disks I: axisymmetric stability

### 2017-01-06 linear.tex
minor edits on SGI stuff

\begin{align}
  \ii\sigma W &= \ii k_x \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi 
% +  \delta\bm{F}\cdot\hat{\bm{r}}
- W F_r - \ii k_x Q - \ii k_x\dd\psi,\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &= - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right] - \dd\psi^\prime  %\delta\bm{F}\cdot\hat{\bm{z}} 
,\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k_x \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2} %, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy}\\
\dd\psi^{\prime\prime}  &= 4\pi G \rho W + k_x^2\dd\psi, \label{lin_sg}
\end{align}  
where $^\prime \equiv \p_z$ and recall $\bm{F} \equiv -\nabla P/\rho$. 
The linearized dust-duffusion function 
$\dd\mathcal{C}$ is given in  Appendix \ref{lin_dust}. 
% and $\dd\bm{F}$ is the linearized pressure
%force, given in Appendix \ref{lin_press}. 
Note that we have assumed a temperature profile that only depends on $r$.

@@

Eq. \ref{lin_mass}---\ref{lin_sg} is a set of ordinary
differntial equations in $z$. All coefficients and amplitudes are
evaluated at a fiducial radius $r=r_0$, but their full $z$-dependence
is retained. We now discuss solutions to these equations for a
razor-thin, self-gravitating disk in \S\ref{sgi}; and for 3D
non-self-gravitating, unstratified and stratified disks 
in \S\ref{si} and \S\ref{results}, respectively.

@@

\section{Secular gravitational instability}\label{sgi}
We first consider a razor-thin disk so that $\rho =
\Sigma\delta(z)$. Here $\delta$ is the delta function and $\Sigma$ is
the total surface density. Note that $\epsilon$ and 
$\tepsilon$ now refers to the global dust-to-gas ratio and dust 
fraction, respectively. We neglect the vertical dimension   
in Eq. \ref{lin_mass}---\ref{lin_energy}. The background disk is
uniform. The linearized dust-gas drag term is then
$-\delta\mathcal{C}/\rho = \tstop\tepsilon c_s^2k_x^2 Q$. The
thin-disk solution to Eq. \ref{lin_sg} is $\dd\psi = -2\pi G
\Sigma W/\left|k_x\right|$

These simplifications yield the dispersion relation
\begin{align*}
  \left(\ii\sigma - \tstop\tepsilon c_s^2k_x^2\right)\left( 2 \pi G
    \Sigma \left|k_x\right|  - \kappa^2 + \sigma^2 \right) = \ii
  \sigma c_s^2 k_x^2\left(1 - \tepsilon\right). 
\end{align*}
Searching for slowly and purely growing modes, $\sigma = \ii s$ with 
$|s|\ll \kappa$, we find 
\begin{align}  
s = \frac{\tstop\tepsilon c_s^2 k_x^2 \left( 2 \pi \Sigma G
    \left|k_x\right| - \kappa^2\right)}{\kappa^2 -2 \pi \Sigma G
    \left|k_x\right| + c_s^2k_x^2\left(1-\tepsilon\right) }. \label{sgi_disp}
\end{align}
This is just the secular gravitational instability in the absence of
turbulent dust diffusion but strong drag
\citep[][ their Eq. 13 becomes Eq. \ref{sgi_disp} in this limit with
a change of variables]{takahashi14}. This is gravitational instability
mediated by dust-gas drag. A similar effect occurs in viscous
self-gravitating gas disks \citep{gammie96,lin16}. In fact, if we
identify $\nu \equiv \tstop c_s^2$ as a kinematic viscosity, then
Eq. \ref{sgi_disp} is effectively 
\citeauthor{gammie96}'s Eq. 18.

This exercise shows that the one-fluid framework, futher simplified by
the terminal velocity approximation, is sufficient to capture the SGI
in the strong drag limit.

\section{Streaming instability}\label{si}
We now consider 3D but non-self-gravitating disks.  
In the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain an algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability (SI) in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

>>>>

\begin{align}
  \ii\sigma W &= \ii k_x \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi 
% +  \delta\bm{F}\cdot\hat{\bm{r}}
- W F_r - \ii k_x Q - \ii k_x\dd\psi,\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &= - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right] - \dd\psi^\prime  %\delta\bm{F}\cdot\hat{\bm{z}} 
,\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k_x \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2} %, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy}\\
\dd\psi^{\prime\prime}  &= 4\pi G \rho W + k_x^2\dd\psi, \label{lin_sg}
\end{align}  
where $^\prime \equiv \p_z$ and recall $\bm{F} \equiv -\nabla
P/\rho$. Self-gravity has been restored temporarily for the next
section (only). The linearized dust-duffusion function 
$\dd\mathcal{C}$ is given in  Appendix \ref{lin_dust}. 
% and $\dd\bm{F}$ is the linearized pressure
%force, given in Appendix \ref{lin_press}. 
Note that we have assumed a temperature profile that only depends on $r$.

@@

Eq. \ref{lin_mass}---\ref{lin_sg} is a set of ordinary
differntial equations in $z$. All coefficients and amplitudes are
evaluated at a fiducial radius $r=r_0$, but their full $z$-dependence
is retained. We next discuss solutions to these equations.  We first
show that the above equations yield the secular
gravitational instability and streaming instability in the strong
drag regime in \S\ref{sgi} and \S\ref{si}, respectively. We then
consider 3D, stratified disks in \S\ref{results} to study the dusty
vertical shear instability.

%for a
%razor-thin, self-gravitating disk in \S\ref{sgi}; and for 3D
%non-self-gravitating, unstratified and stratified disks 
%in \S\ref{si} and \S\ref{results}, respectively.

@@

%\section{Connection to previous results}

\subsection{Secular gravitational instability}\label{sgi}
We first consider a razor-thin, self-gravitating disk so that $\rho =
\Sigma\delta(z)$. Here $\delta$ is the Dirac delta function and $\Sigma$ is
the total surface density. Now $\epsilon$ and 
$\tepsilon$ refers to the global dust-to-gas ratio and dust 
fraction, respectively. The background disk is
uniform and we neglect the vertical dimension   
in Eq. \ref{lin_mass}---\ref{lin_energy}. The linearized dust-gas drag
term is then $-\delta\mathcal{C}/\rho = \tstop\tepsilon c_s^2k_x^2
Q$. The thin-disk solution to Eq. \ref{lin_sg} is $\dd\psi = -2\pi G
\Sigma W/\left|k_x\right|$

These simplifications yield the dispersion relation
\begin{align*}
  \left(\ii\sigma - \tstop\tepsilon c_s^2k_x^2\right)\left( 2 \pi G
    \Sigma \left|k_x\right|  - \kappa^2 + \sigma^2 \right) = \ii
  \sigma c_s^2 k_x^2\left(1 - \tepsilon\right). 
\end{align*}
Searching for slowly and purely growing modes, $\sigma = \ii s$ with 
$|s|\ll \kappa$, we find 
\begin{align}  
s = \frac{\tstop\tepsilon c_s^2 k_x^2 \left( 2 \pi \Sigma G
    \left|k_x\right| - \kappa^2\right)}{\kappa^2 -2 \pi \Sigma G
    \left|k_x\right| + c_s^2k_x^2\left(1-\tepsilon\right) }. \label{sgi_disp}
\end{align}
This is secular gravitational instability mediated by strong
dust-gas drag with negligible turbulent dust diffusion 
\citep[][ their Eq. 13 becomes our Eq. \ref{sgi_disp} in this limit
after 
a change of variables]{takahashi14}. A similar effect occurs in viscous
self-gravitating gas disks \citep{gammie96,lin16}. In fact, if we
identify $\nu \equiv \tstop \tepsilon c_s^2$ as a kinematic viscosity,
then Eq. \ref{sgi_disp} is identical to \citeauthor{gammie96}'s Eq. 18.

This exercise shows that the one-fluid framework, futher simplified by
the terminal velocity approximation, is sufficient to capture SGI
in the strong drag limit.

\subsection{Streaming instability}\label{si}
We now consider 3D but non-self-gravitating disks.  
In the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain an algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability (SI) in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

### 2017-01-06 model.tex
minor edits on SGI stuff

The total gravitational potential $\Phi_\mathrm{tot}=\Phi + \psi$,
where $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
and $\psi$ is the disk potential satisfying
\begin{align}
  \nabla^2\psi = 4 \pi G \rho.
\end{align}
Here, $G$ is the gravitational constant, and $(r,\phi, z)$ are 
cylindrical co-ordinates centered on the star.  
The second equality in Eq. \ref{thin_disk_potential} is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{eqm}).

We include the gravitational potential of the mixture 
$\psi$ for completeness but will neglect it in most of this
work. Henceforth we consider non-self-gravitating disks ($\psi \equiv
0$) unless explicitly stated otherwise.

@@

@@

The physical interpretation of Eq. \ref{thermal_instability} is that 
work done by pressure forces in the dusty gas leads to growth in
oscillation amplitudes. To demonstrate this, we calculate the average
work done during a periodic oscillation, and show that if the work
done is positive, then the 
oscillation amplitude would actually grow.

>>>>

The total gravitational potential is $\Phi_\mathrm{tot}=\Phi + \psi$,
where $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
and $\psi$ is the disk potential satisfying
\begin{align}
  \nabla^2\psi = 4 \pi G \rho.
\end{align}
Here, $G$ is the gravitational constant, and $(r,\phi, z)$ are 
cylindrical co-ordinates centered on the star.  
The second equality in Eq. \ref{thin_disk_potential} is the 
thin-disk approximation for the stellar potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{eqm}).

We include the gravitational potential of the mixture 
$\psi$ for completeness but will neglect it throughout most of this
work ($\psi \equiv 0$) unless explicitly stated otherwise.

@@

{\bf
\subsection{Physical disk conditions}
physical disk conditions underwhich the above approx
(small/well-coupled dust and isothermal gas) is applicable. smaller
than mm at few tens of au. give dimensional numbers here
}

@@

The physical interpretation of Eq. \ref{thermal_instability} is that 
work done by pressure forces in the dusty gas leads to growth in
oscillation amplitudes. To demonstrate this, we calculate the average
work done assuming periodic oscillations, and show that if the average
work done is positive, then the 
oscillation amplitude would actually grow.

### 2017-01-06 paper.tex
minor edits on SGI stuff

We begin our formal discussion in \S\ref{setup} by listing the 
basic equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} that are used in
explicit calculations. In \S\ref{limits} we discuss 
general stability properties of dusty disks. We then move onto
explicit stability analyses. In  
\S\ref{linear_problem} we list the linearized equations. We solve them 
to recover SGI in \S\ref{sgi}, to interpret the SI in \S\ref{si}, and
in  \S\ref{results} to extend the VSI to dusty disks. We  
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

@@

From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure (i.e. restoring) forces is still 
increasing. This over-compensation causes the parcel to expand
beyond the maximum volume of the previous cycle. 
Similarly, from $C$ to $D$ the parcel is already contracting towards
equilibrium, but pressure 
forces are still dropping. This allows the  
parcel's contraction to over-shoot the maximum density attained in the
previous cycle. 
The overall positive work done leads to growth in the oscillation amplitude. 
This effect is similar to being pushed downwards on a swing on
descent.

@@

There are several situations where the pressure and density of a fluid
are not in phase. The obvious case is if the fluid is subject to
external heating/cooling. For example, in strongly irradiated
protoplanetary disks the
disk temperature  $T(r)$ is time-independent but varies with radius
\citep{chiang97}. Since $P\propto \rho T$ for an ideal gas, the pressure of a fluid parcel varies with 
position as well. There is no reason to expect $P$ and $\rho$ to
be in phase. In fact, we will show that this is responsible for 
the vertical shear instability \citep{lin15}.

>>>>

We begin our formal discussion in \S\ref{setup} by listing the 
basic equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} that are used in
explicit calculations. In \S\ref{limits} we discuss 
general stability properties of dusty disks. We then move onto
explicit stability analyses. In  
\S\ref{linear_problem} we list the linearized equations and show that 
they reproduce previous results on SGI and SI. In \S\ref{results} we
extend the VSI to dusty disks. We   
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

@@

From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure (i.e. restoring) forces is still 
increasing. This over-compensation causes the parcel to expand
beyond the maximum volume of the previous cycle. 
Similarly, from $C$ to $D$ the parcel is already contracting towards
equilibrium, but pressure 
forces are still dropping. This allows the  
parcel's contraction to over-shoot the maximum density attained in the
previous cycle. 
The overall positive work done leads to growth in the oscillation amplitude. 
This effect is similar to being pushed downwards on a swing when 
descending.

@@

There are several situations where the pressure and density of a fluid
are not in phase. The obvious case is if the fluid is subject to
external heating/cooling. For example, in strongly irradiated
protoplanetary disks the
disk temperature  $T(r)$ is time-independent but varies with radius
$r$ from the star \citep{chiang97}. Since $P\propto \rho T$ for an ideal gas, the pressure of a fluid parcel varies with 
position as well. There is no reason to expect $P$ and $\rho$ to
be in phase. In fact, we will show that this is responsible for 
the vertical shear instability \citep{lin15}.
